---
title: "Querying a MySQL database"
format:  html
  
execute: 
  echo: true
editor_options: 
  chunk_output_type: inline
---


```{r global_options, include=FALSE}
posted <- Sys.time() > "2020-11-25 11:00:00 EDT"
```


```{r, message=FALSE, eval=FALSE}
# do NOT install RMySQL if you are on the RStudio Server
install.packages("RMySQL")
```



```{r, message=FALSE}
library(tidyverse)
library(RMySQL)
db <- dbConnect(
  MySQL(),
  host = "scidb.smith.edu",
  user = "sds192",
  password = "DSismfc@S",
  dbname = "imdb"
  )
knitr::opts_chunk$set(connection= 'db', max.print = 5)
```

 

```{r}
class(db)
```



```{r}

dbGetQuery(conn = db,
           "SHOW DATABASES") |>
  head(5)
```



```{r}
dbGetQuery(conn = db,"USE imdb;")
```



```{r}
dbGetQuery(conn = db, "SHOW TABLES;")
```


```{r}
#| output = TRUE
dbGetQuery(conn = db, "SELECT * FROM kind_type;")
```



```{r}
## output.var="info_types" in this chunk

info_types<- dbGetQuery(conn = db, "SELECT * FROM info_type;")
```

.

```{r}
# Note: this is an r chunk s we use R instead of SQL.

head(info_types)
```






```{r}
dbGetQuery(conn = db, 
           "SELECT *
           FROM title
           WHERE title= 'Groundhog Day'
           LIMIT 10;")
```



```{r}
dbGetQuery(conn = db, "SELECT *
           FROM title
           WHERE title = 'Groundhog Day'
           AND kind_id = 1;")
```



```{r}
dbGetQuery(conn = db, "SELECT *
           FROM title
           WHERE title LIKE '%Groundhog%'
           AND kind_id = 1;")
```


```{r}
dbGetQuery(conn = db, "SELECT *
           FROM title
           WHERE title LIKE '%Groundhog%'
           AND kind_id = 1
           ORDER BY production_year;")
```



```{r}
dbGetQuery(conn = db, "SELECT t.id, t.title, t.production_year
           FROM title AS t
           WHERE title LIKE '%Groundhog%'
           AND t.kind_id = 1
           ORDER BY t.production_year;")
```



```{r}
dbGetQuery(conn = db, "SELECT *
           FROM title
           WHERE title LIKE '%Ghostbusters%'
           AND kind_id = 1
           LIMIT 10;")
  

```


```{r}
dbGetQuery(conn= db, "DESCRIBE name;")
```




```{r}
dbGetQuery(conn= db,"SELECT id
           FROM name
           WHERE name LIKE 'MacDowell, Andie%';")

```




```{r}
dbGetQuery(conn= db, "SELECT *
           FROM cast_info
           WHERE movie_id = 3664274
           LIMIT 10;")
```


I've learned a lot from working with SQL because it's a language that lets me communicate directly with databases, which opens up a ton of data for analysis. For those working with data, SQL is a vital tool due to its ease of use and strong database searching capabilities. Each search provides information about the underlying data structure and relationships, ranging from straightforward SELECT queries to challenging JOINs and groups. One of my favorite tools in the data analysis toolset is SQL because of its capacity to handle data at scale and extract valuable information. Discovering patterns, trends, and correlations that inform decisions and improve comprehension is just as important as just analyzing data.